Forecasting the 2024 US Presidential Election: A Poll-Based Approach

### 2024-10-31 paper/paper.qmd
Updated primary model formulation

library(tidyverse)
library(skimr)
library(knitr)
library(GGally)
library(dplyr)
```

@@

## Model set-up
```{r}
#| echo: false
#| message: false
#| warning: false
data_nevada <- data %>% filter(state == "Nevada") 
data_arizona <- data %>% filter(state == "Arizona") 
data_wisconsin <- data %>% filter(state == "Wisconsin") 
data_michigan <- data %>% filter(state == "Michigan") 
data_penn <- data %>% filter(state == "Pennsylvania") 
data_north_carolina <- data %>% filter(state == "North Carolina") 
data_georgia <- data %>% filter(state == "Georgia")

candidate_names <- c("Kamala Harris", "Donald Trump")

fit_model <- function(data, candidate) {
  candidate_data <- data %>% filter(candidate_name == candidate)

@@

@@

```{r}
intercept_results <- final_results %>%
  filter(term == "(Intercept)")

@@

```{r}

The linear regression model is specified for each candidate separately, with the following structure:

\begin{align}
\text{pct}_i & = \beta_0 + \beta_1 \text{pollscore}_i + \beta_2 \text{transparency\_score}_i + \beta_3 \text{sample\_size}_i + \beta_4 \text{state}_i + \beta_5 \text{days\_towards\_election}_i + \epsilon_i \\
\epsilon_i & \sim \text{Normal}(0, \sigma^2)
\end{align}

Where:
\begin{align}
\beta_0 & \text{ is the intercept term} \\
\beta_1, \beta_2, \beta_3, \beta_4, \beta_5 & \text{ are the coefficients for each predictor} \\
\sigma^2 & \text{ is the variance of the error term}
\end{align}

In the implementation, a function is defined to fit this model for each candidate (Kamala Harris and Donald Trump) individually. The model fits are then summarized to interpret the coefficients for each predictor.

>>>>

library(tidyverse)
library(skimr)
library(knitr)
library(GGally)
library(dplyr)
library(broom)
```

@@

## Model set-up

```{r baseline model}
#| echo: false
#| message: false
#| warning: false
fit_model <- function(data, candidate) {
  candidate_data <- data %>% filter(candidate_name == candidate)

@@

# Candidate names
candidate_names <- c("Kamala Harris", "Donald Trump")

# Loop over candidates and store each model's results
for (candidate in candidate_names) {
  model_result <- fit_model(data, candidate)  # Fit the model

  # Check if the model_result is not NULL and add to the results list
  if (!is.null(model_result)) {
    results[[candidate]] <- model_result
  }
}

```{r our model}
#| echo: false
#| message: false
#| warning: false
data_nevada <- data %>% filter(state == "Nevada") 
data_arizona <- data %>% filter(state == "Arizona") 
data_wisconsin <- data %>% filter(state == "Wisconsin") 
data_michigan <- data %>% filter(state == "Michigan") 
data_penn <- data %>% filter(state == "Pennsylvania") 
data_north_carolina <- data %>% filter(state == "North Carolina") 
data_georgia <- data %>% filter(state == "Georgia")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
intercept_results <- final_results %>%
  filter(term == "(Intercept)")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
solid_state_votes <- data.frame(
  Candidate = c("Harris", "Trump"),
  Solid_State_Votes = c(226, 219)
)
intercept_wide$electoral_votes <- c(6, 11, 10, 15, 19, 16, 16)

swing_state_results <- intercept_wide %>%
  mutate(Winner = ifelse(difference > 0, "Harris", "Trump")) %>%
  group_by(Winner) %>%
  summarize(Predicted_Swing_State_Votes = sum(electoral_votes))

total_votes <- solid_state_votes %>%
  left_join(swing_state_results, by = c("Candidate" = "Winner")) %>%
  mutate(Predicted_Swing_State_Votes = ifelse(is.na(Predicted_Swing_State_Votes), 0, Predicted_Swing_State_Votes),
         Total_Predicted_Votes = Solid_State_Votes + Predicted_Swing_State_Votes)

kable(total_votes, caption = "Predicted Electoral Votes for Each Candidate")
```

### Model Formulations

#### 1. Baseline Model (Winner-Take-All National Model)
The baseline model aggregates national polling data to predict a winner, with all electoral votes awarded to the candidate with the higher national polling percentage.

1. **National Polling Projections**  
   Let:
   - $P_H(t)$: Harris’s national polling percentage at $t$ days towards the election.
   - $P_T(t)$: Trump’s national polling percentage at $t$ days towards the election.

   We model each candidate’s polling as:
   $$
   P_H(t) = \alpha_H + \beta_H t
   $$
   $$
   P_T(t) = \alpha_T + \beta_T t
   $$

   where $\alpha_H$ and $\alpha_T$ are the intercepts (predicted percentages on election day, $t = 0$), and $\beta_H$ and $\beta_T$ are the slopes.

2. **Election Day Prediction**  
   On election day:
   $$
   P_H(0) = \alpha_H, \quad P_T(0) = \alpha_T
   $$

3. **Winner-Take-All Outcome**  
   The candidate with the higher predicted national percentage wins all electoral votes:
   $$
   W = 
   \begin{cases} 
   \text{Harris wins (all electoral votes)} & \text{if } \alpha_H > \alpha_T \\
   \text{Trump wins (all electoral votes)} & \text{otherwise}
   \end{cases}
   $$

#### 2. Primary Model (Swing State-Based Model)
This model forecasts electoral outcomes by projecting state-level polling results for each candidate, combining swing state outcomes with votes from solid states.

1. **Swing State Polling Projections**  
   For each swing state $s \in S$, let:
   - $P_{H_s}(t)$: Harris’s polling percentage in state $s$ at $t$ days before the election.
   - $P_{T_s}(t)$: Trump’s polling percentage in state $s$ at $t$ days before the election.

   We model each swing state polling as:
   $$
   P_{H_s}(t) = \alpha_{H_s} + \beta_{H_s} t
   $$
   $$
   P_{T_s}(t) = \alpha_{T_s} + \beta_{T_s} t
   $$

   where $\alpha_{H_s}$ and $\alpha_{T_s}$ are intercepts (predicted percentages at $t = 0$) and $\beta_{H_s}$, $\beta_{T_s}$ are slopes.

2. **Predicting Swing State Winners**  
   On election day ($t = 0$), each candidate’s predicted support in state $s$is:
   $$
   P_{H_s}(0) = \alpha_{H_s}, \quad P_{T_s}(0) = \alpha_{T_s}
   $$

   Define $W_s$ as an indicator for the swing state winner:
   $$
   W_s = 
   \begin{cases} 
   1 & \text{if } \alpha_{H_s} > \alpha_{T_s} \\
   0 & \text{otherwise}
   \end{cases}
   $$

3. **Electoral Vote Aggregation**  
   Let $EV_H$ and $EV_T$ represent total electoral votes for Harris and Trump:
   $$
   EV_H = \sum_{s \in S} W_s \cdot EV_s + \sum_{s \in S_H} EV_s
   $$
   $$
   EV_T = \sum_{s \in S} (1 - W_s) \cdot EV_s + \sum_{s \in S_T} EV_s
   $$

4. **Final Election Outcome**  
   The candidate surpassing 270 electoral votes wins:
   $$
   W = 
   \begin{cases} 
   \text{Harris wins} & \text{if } EV_H \geq 270 \\
   \text{Trump wins} & \text{if } EV_T \geq 270 
   \end{cases}
   $$
